A memory frontier for complex synapses

### 2013-08-27 NIPS/SynapseNIPS.tex
typo in poster, removed repeated fig in NIPS paper

The extent to which the memory has been stored is described by a signal-to-noise ratio (SNR) \cite{Fusi2005cascade,Fusi2007multistate,Barrett2008discrete}:
%
\begin{equation}\label{eq:SNRdef}
  \snr(t) = \frac{\av{\wv_\text{ideal}\cdt\wv(t)} - \av{\wv_\text{ideal}\cdt\wv(\infty)}}
     {\sqrt{\var(\wv_\text{ideal}\cdt\wv(\infty))}}.
\end{equation}
%
The noise in the denominator is essentially $\sqrt{N}$.
There is a correction when potentiation and depression are imbalanced, but this will not affect the upper bounds that we will discuss below and will be ignored in the subsequent formulae.

@@

With the states in the order described above, we can find perturbations of $\M\potdep$ that will always increase the area, whilst leaving the equilibrium distribution, $\eq$, unchanged.
Some of these perturbations are shown in \autoref{fig:perts}.
For example, in \autoref{fig:perts}\ref{fig:tri_right}, for two states $i$ on the left and $j$ on the right, with $j$ being more ``potentiated" than $i$ (i.e. $\eta^+_i > \eta^+_j$), we have proven that increasing $\M\pot_{ij}$ and decreasing $\M\dep_{ij}$ leads to an increase in area.
The only thing that can prevent these perturbations from increasing the area is when they require the decrease of a matrix element that has already been set to 0.
This determines the topology (non-zero transition probabilities) of the model with maximal area.
It is of the form shown in \autoref{fig:max_area}, with potentiation moving one step to the right and depression moving one step to the left.
Any other topology would allow some perturbation to further increase the area.

@@

As these perturbations do not change the equilibrium distribution, this means that the area of \emph{any} model is bounded by that of a linear chain with the same equilibrium distribution.
The area of a linear chain model can be expressed directly in terms of its equilibrium state distribution, $\eq$, yielding the following upper bound on the area of any model with the same $\eq$ (see supplementary material):
%
\begin{equation}\label{eq:multistatearea}
  A \leq \frac{2\sqrt{N}}{r} \sum_k \brk{k - \sum_j j\eq_j} \eq_k \w_k
    = \frac{2\sqrt{N}}{r} \sum_k \abs{k - \sum_j j\eq_j} \eq_k ,
\end{equation}
%
where we chose $\w_k=\sgn\brk{k - \sum_j j\eq_j}$.
We can then maximize this by pushing all of the equilibrium distribution symmetrically to the two end states.
This can be done by reducing the transition probabilities out of these states, as in \autoref{fig:max_area}.
This makes it very difficult to get out of the end states once they have been entered.
The resulting area is
%
\begin{equation}\label{eq:max_area}
  A \leq \frac{\sqrt{N}(M-1)}{r}.
\end{equation}
%
This analytical result is similar to a numerical result found in \cite{Barrett2008discrete} under a slightly different information theoretic measure of memory performance.

The ``sticky'' end states result in very slow decay of memory, but they also make it difficult to encode the memory in the first place, since a small fraction of synapses are able to change synaptic efficacy during the storage of a new memory.
Thus models that optimize area optimize memory at late times, at the expense of early times.

\begin{figure}[tbp]
 \begin{center}
  \includegraphics[width=0.35\linewidth]{multistate_sticky.svg}
 \end{center}
  \caption{Model that maximizes the area.
  Unlabeled transitions have probability 1.
  Labeled transitions have probability $\varepsilon\to0$.
  }\label{fig:max_area}
\end{figure}

@@

We see that at early, but not late times, there is a gap between the upper bound that we can prove and what we can achieve with actual models.
There may be other models we haven't found that could beat the ones we have, and come closer to our proven envelope.
However, we suspect that the area constraint is not the bottleneck for optimizing memory at times less than  $\CO(\frac{M}{r})$.
We believe there is some other constraint that prevents models from approaching the envelope, and currently are exploring several mathematical conjectures for the precise form of this constraint, in order to obtain a potentially tighter envelope.
Nevertheless, we have proven rigorously that no model's memory curve can ever exceed this envelope, and that it is at least tight for late times, longer than $\CO(\frac{M}{r})$, where models of the form in \autoref{fig:max_area} can come close to the envelope.

\begin{figure}
 \begin{center}
 \begin{myenuma}
 \parbox[t]{0.45\linewidth}{
  \item\aligntop{\includegraphics[width=0.95\linewidth]{env.svg}}\label{fig:envplot}
  }\hp
  \parbox[t]{0.45\linewidth}{
  \item\aligntop{\includegraphics[width=0.8\linewidth]{diffjump.svg}}\label{fig:diffjump}\\[0.1\linewidth]
  \item\aligntop{\includegraphics[width=0.8\linewidth]{multistate_sticky.svg}}\label{fig:sticky}
  }
 \end{myenuma}
 \end{center}
  \caption{The memory curve envelope for $N=100$, $M=12$.
  (\ref{fig:envplot}) An upper bound on the SNR at any time is shown in green.
  The red dashed curve shows the result of numerical optimization of synaptic models with random initialization.
  The solid red curve shows the highest SNR we have found with hand designed models.
  At early times these models are of the form shown in (\ref{fig:diffjump}) with different numbers of states, and all transition probabilities equal to $1$.
  At late times they are of the form shown in (\ref{fig:sticky}) with different values of $\varepsilon$.
  }\label{fig:env}
\end{figure}

>>>>

The extent to which the memory has been stored is described by a signal-to-noise ratio (SNR) \cite{Fusi2005cascade,Fusi2007multistate}:
%
\begin{equation}\label{eq:SNRdef}
  \snr(t) = \frac{\av{\wv_\text{ideal}\cdt\wv(t)} - \av{\wv_\text{ideal}\cdt\wv(\infty)}}
     {\sqrt{\var(\wv_\text{ideal}\cdt\wv(\infty))}}.
\end{equation}
%
The noise in the denominator is essentially $\sqrt{N}$.
There is a correction when potentiation and depression are imbalanced, but this will not affect the upper bounds that we will discuss below and will be ignored in the subsequent formulae.

@@

With the states in the order described above, we can find perturbations of $\M\potdep$ that will always increase the area, whilst leaving the equilibrium distribution, $\eq$, unchanged.
Some of these perturbations are shown in \autoref{fig:perts}, see supplementary material for details.
For example, in \autoref{fig:perts}(\ref{fig:tri_right}), for two states $i$ on the left and $j$ on the right, with $j$ being more ``potentiated" than $i$ (i.e. $\eta^+_i > \eta^+_j$), we have proven that increasing $\M\pot_{ij}$ and decreasing $\M\dep_{ij}$ leads to an increase in area.
The only thing that can prevent these perturbations from increasing the area is when they require the decrease of a matrix element that has already been set to 0.
This determines the topology (non-zero transition probabilities) of the model with maximal area.
It is of the form shown in \autoref{fig:env}(\ref{fig:sticky}),%\autoref{fig:max_area}, 
with potentiation moving one step to the right and depression moving one step to the left.
Any other topology would allow some perturbation to further increase the area.

@@

As these perturbations do not change the equilibrium distribution, this means that the area of \emph{any} model is bounded by that of a linear chain with the same equilibrium distribution.
The area of a linear chain model can be expressed directly in terms of its equilibrium state distribution, $\eq$, yielding the following upper bound on the area of any model with the same $\eq$ (see supplementary material):
%
\begin{equation}\label{eq:multistatearea}
  A \leq \frac{2\sqrt{N}}{r} \sum_k \brk{k - \sum_j j\eq_j} \eq_k \w_k
    = \frac{2\sqrt{N}}{r} \sum_k \abs{k - \sum_j j\eq_j} \eq_k ,
\end{equation}
%
where we chose $\w_k=\sgn\brk{k - \sum_j j\eq_j}$.
We can then maximize this by pushing all of the equilibrium distribution symmetrically to the two end states.
This can be done by reducing the transition probabilities out of these states, as in \autoref{fig:env}(\ref{fig:sticky}).%\autoref{fig:max_area}.
This makes it very difficult to get out of the end states once they have been entered.
The resulting area is
%
\begin{equation}\label{eq:max_area}
  A \leq \frac{\sqrt{N}(M-1)}{r}.
\end{equation}
%
This analytical result is similar to a numerical result found in \cite{Barrett2008discrete} under a slightly different information theoretic measure of memory performance.

The ``sticky'' end states result in very slow decay of memory, but they also make it difficult to encode the memory in the first place, since a small fraction of synapses are able to change synaptic efficacy during the storage of a new memory.
Thus models that maximize area optimize memory at late times, at the expense of early times.

%\begin{figure}[tbp]
% \begin{center}
%  \includegraphics[width=0.35\linewidth]{multistate_sticky.svg}
% \end{center}
%  \caption{Model that maximizes the area.
%  Unlabeled transitions have probability 1.
%  Labeled transitions have probability $\varepsilon\to0$.
%  }\label{fig:max_area}
%\end{figure}

@@

We see that at early, but not late times, there is a gap between the upper bound that we can prove and what we can achieve with actual models.
There may be other models we haven't found that could beat the ones we have, and come closer to our proven envelope.
However, we suspect that the area constraint is not the bottleneck for optimizing memory at times less than  $\CO(\frac{M}{r})$.
We believe there is some other constraint that prevents models from approaching the envelope, and currently are exploring several mathematical conjectures for the precise form of this constraint in order to obtain a potentially tighter envelope.
Nevertheless, we have proven rigorously that no model's memory curve can ever exceed this envelope, and that it is at least tight for late times, longer than $\CO(\frac{M}{r})$, where models of the form in \autoref{fig:env}(\ref{fig:sticky})%\autoref{fig:max_area}
 can come close to the envelope.

\begin{figure}
 \begin{center}
 \begin{myenuma}
 \parbox[t]{0.45\linewidth}{
  \item\aligntop{\includegraphics[width=0.95\linewidth]{env.svg}}\label{fig:envplot}
  }\hp
  \parbox[t]{0.45\linewidth}{
  \item\aligntop{\includegraphics[width=0.8\linewidth]{diffjump.svg}}\label{fig:diffjump}\\[0.1\linewidth]
  \item\aligntop{\includegraphics[width=0.8\linewidth]{multistate_sticky.svg}}\label{fig:sticky}
  }
 \end{myenuma}
 \end{center}
  \caption{The memory curve envelope for $N=100$, $M=12$.
  (\ref{fig:envplot}) An upper bound on the SNR at any time is shown in green.
  The red dashed curve shows the result of numerical optimization of synaptic models with random initialization.
  The solid red curve shows the highest SNR we have found with hand designed models.
  At early times these models are of the form shown in (\ref{fig:diffjump}) with different numbers of states, and all transition probabilities equal to $1$.
  At late times they are of the form shown in (\ref{fig:sticky}) with different values of $\varepsilon$.
  The model shown in (\ref{fig:sticky}) also saturates the area bound \eqref{eq:max_area} in the limit $\varepsilon\to0$.
  }\label{fig:env}
\end{figure}
